Graphical Tests for Power Comparison of Competing Designs

### 2012-03-29 setup.tex
table formatting

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different". They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This study allows us to make informed judgments about how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, participants were shown two specially prepared lineups as a baseline of performance -- if participants got just one of these two reference lineups correct, we required at least an overall correct percentage of 20\%, which is indicative of a performance that is better than guessing. Table \ref{tbl:treatment} gives an overview of the number of times lineups in each combination of design and sample size were shown, and in how many of them the data plot was correctly identified.

@@

For evaluating the results, we are using a linear mixed effects modeling \cite{pinheiro:2000} approach in all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}.

@@

Overall discussion: huge difference in power between polar charts and Euclidean, reference line has surprisingly little influence, but more so for polar charts than for barcharts,
An increase in sample size has positive impact on correctness (see figure \ref{fig:treatment} for an overview). Polar charts need much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The offset changes are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. Polar charts benefit more from this inversion than barcharts. This is an unexpected finding, but is consistent throughout different lineups in the data.

@@

\begin{table}[ht]
\begin{center}
%\resizebox{\linewidth}{!} {
\rowcolors{2}{white}{lightgray}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Error & $z$-value & $p$-value \\ 
  \hline
euclid & -0.08 & 0.39 & -0.21 & 0.84 \\ 
polar & -1.98 & 0.32 & -6.13 & 0.00 \\ [1pt]
  reflineTRUE & -0.14 & 0.26 & -0.53 & 0.59 \\ [1pt]
  sample\_size & 0.27 & 0.04 & 6.31 & 0.00 \\ [1pt]
  offset 90 & -0.43 & 0.37 & -1.18 & 0.24 \\ 
  offset 180 & -0.89 & 0.35 & -2.51 & 0.01 \\ 
  offset 270 & 0.21 & 0.38 & 0.55 & 0.58 \\ [1pt]
  polar:reflineTRUE & 0.51 & 0.35 & 1.44 & 0.15 \\ [1pt]
  polar:sample\_size & -0.23 & 0.05 & -5.02 & 0.00 \\ [1pt]
  polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 \\ 
  polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 \\
   \hline
\end{tabular}
%}
\end{center}
\caption{\label{tbl:correct} Model output for correct result. }
\end{table}

>>>>

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different". They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This study allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, participants were shown two specially prepared lineups as a baseline of performance -- if participants answered just one of these two reference lineups correctly, we required at least an overall correct percentage of 20\%, which is indicative of a performance that is better than guessing. Table \ref{tbl:treatment} gives an overview of the number of times lineups in each combination of design and sample size were shown, and in how many of them the data plot was correctly identified.

@@

For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates of {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and as two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. The reference line has surprisingly little influence, but it helps more for  polar charts than for barcharts,
An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview). Polar charts need much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The offset changes are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. Polar charts benefit more from this inversion than barcharts. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the 'mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

@@

\begin{table}[ht]
\begin{center}
\resizebox{\linewidth}{!} {
\rowcolors{2}{white}{lightgray}
\begin{tabular}{rrrrrrl}
  \hline
& & Estimate & Error & $z$-value & $p$-value &\\ 
  \hline
design & euclid & -0.08 & 0.39 & -0.21 & 0.84 &\\ 
&polar & -1.98 & 0.32 & -6.13 & 0.00 & ***\\ [3pt]
main effects &&&&&&\\
reference & line  & -0.14 & 0.26 & -0.53 & 0.59 & \\ [1pt]
size &  sample size & 0.27 & 0.04 & 6.31 & 0.00 & ***\\[1pt]
offset &  90 degrees& -0.43 & 0.37 & -1.18 & 0.24 &\\ 
  & 180 degrees& -0.89 & 0.35 & -2.51 & 0.01 & **\\ 
  & 270 degrees& 0.21 & 0.38 & 0.55 & 0.58 &\\ [3pt]
interactions  &&&&&&\\
&  polar:line & 0.51 & 0.35 & 1.44 & 0.15 &\\ [1pt]
&    polar:sample size & -0.23 & 0.05 & -5.02 & 0.00 & ***\\[1pt]
&    polar:offset 90 & 0.64 & 0.49 & 1.30 & 0.20 \\ 
&  polar:offset 180 & 0.91 & 0.47 & 1.92 & 0.05 & .\\ 
&    polar:offset 270 & -0.73 & 0.54 & -1.35 & 0.18 &\\
   \hline
\end{tabular}
}
\end{center}
\caption{\label{tbl:correct} Output of a generalized linear mixed effects model for power of lineups (i.e. probability of identifying the data plot) for comparing designs. Included are two-way interactions with sample size and shifts in wind direction (offset). Results are based on  976 lineup evaluations by 115 participants. }
\end{table}
